$p$-adic precision

### 2014-02-06 padicprec.tex
Differentials of gcd's

\subsection{Univariate polynomials}

@@

\todo{To be written.}

@@

@@

We can develop further this example and study what happens on the 
closed subset of $M_{n,n-d}(K)$ where matrices have not full rank. On
this subspace, the left kernel has dimension $< d$ and then no longer
defines a point in $\Grass(n,d)$. Nevertheless, for all integer $r <
n-d$, we can consider the subset $\mathcal V_r \subset M_{n,n-d}(K)$
of matrices whose rank are exactly rank $r$. It is locally closed in
$M_{n,n-d}(K)$ with respect to the Zariski topology and hence defines
a $K$-manifold. Furthermore, we have a mapping $\text{LK}_r : \mathcal 
M_r \to \Grass(n,n-r)$ which is differentiable and whose differential 
can be computed as before.

The lesson from this study is the following. If we have to compute the 
kernel of a matrix $M$ over $K$ known with finite precision, \emph{a 
priori} we first need to determine what is the rank of $M$. However, the 
rank function is not continuous --- it is only lower semi-continuous --- 
and hence cannot be determined from sure from $M$ (expect if $M$ has 
full rank). We then need to make an extra hypothesis and the more natural 
one is arguably to assume that $M$ has actually the minimal rank that it 
can have. With this convention, for instance, the kernel of the zero 
matrix $M \in M_n(K)$ \emph{known with finite precision} is always the 
whole space although there exist invertible matrices in any 
neighborhood of $M$.

>>>>

\subsection{Univariate polynomials}
\label{ssec:polynomials}

@@

We fix two positive integers $n$ and $m$ with $n \geq m$. We consider the 
function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
couple $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
coefficients of minimal degrees (they are those computed by extended 
Euclide algorithm).
The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
subset $\mathcal V_0$ where the function $\gcd$ takes the constant value 
$1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
of positive degree. This implies that $f$ is not continuous outside 
$\mathcal V_0$. On the contrary, on $\mathcal V_0$, the function $f$ is
differentiable, and actually locally analytic.

Of course, on $\mathcal V_0$, the first component $D$ is constant and 
therefore $dD = 0$. To compute $dU$ and $dV$, one can simply 
differentiate the B\'ezout relation $AU + BV = 1$. We get:
$$A \cdot dU + B \cdot dV = - (dA \cdot U + dB \cdot V)$$
from which we deduce that $dU$ (resp. $dV$) is obtained as the 
remainder in the Euclidean division of $U{\cdot}dX$ by $B$ (resp. of 
$V{\cdot}dX$ by $A$) where $dX = - (dA \cdot U + dB \cdot V)$.

In order to go further and differentiate $f$ outside $\mathcal V_0$, we 
define, for all integer $i$, the subset $\mathcal V_i$ of $K_n[X] \times 
K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
subresultants shows that $\mathcal V_i$ is locally closed with respect to 
the Zariski topology. In particular, it defines a $K$-manifold in the 
sense of Appendix \ref{sec:manifold}. This is interesting because the 
restriction of $f$ to $\mathcal V_i$ is now differentiable. To compute
its differential, we proceed along the same lines as before: we 
differentiate the relation $AU + BV = D$ and obtain this way
$A \cdot dU + B \cdot dV - dD = dX$
with $dX = - (dA \cdot U + dB \cdot V)$. In the above relation, the
first two terms $A{\cdot}dU$ and $B{\cdot}dV$ are divisible by $D$ 
whereas the term $dD$ has degree less than $i$. 
Hence if $dX = D \cdot dQ + dR$ is the Euclidean division of $dX$ by $D$, 
we must have $\frac A D \cdot dU + \frac B D \cdot dV = dQ$ and $dD = 
-dR$. These relations, together with bounds on the degree of $U$ and $V$,
imply as before that $dU$ (resp. $dV$) are equal to the remainder in the
Euclidean division of $U{\cdot}dQ$ by $\frac B D$ (resp. of $V{\cdot}dQ$
by $\frac A D$).

The lesson we may retain from this study is the following. If we have to 
compute the greatest common divisor of two polynomials $A$ and $B$ known 
with finite precision, we first need to determine what is its degree. 
However, the degree function is not continuous --- it is only upper 
semi-continuous --- and hence cannot be determined from sure from $A$ and 
$B$ (expect if they are coprime). We then need to make an extra 
hypothesis and the more natural one is arguably to assume that 
$\gcd(A,B)$ has actually the maximal degree that it can have. Once this
assumption is made, the computation is possible and one can apply Lemma 
\ref{lem:main} to determine the precision of the result.
Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
itself although there exist pair of coprime polynomials in any 
neighborhood of $(A,A)$.

@@

\todo{Discuss a bit multiple roots.}

@@

Following what we have already done in \S \ref{ssec:polynomials} for 
greatest common divisors of polynomials, we can develop further this 
example and study what happens on the closed subset of $M_{n,n-d}(K)$ 
where matrices have not full rank. On this subspace, the left kernel has 
dimension $< d$ and then no longer defines a point in $\Grass(n,d)$. 
Nevertheless, for all integer $r < n-d$, we can consider the subset 
$\mathcal V_r \subset M_{n,n-d}(K)$ of matrices whose rank are exactly 
rank $r$. It is locally closed in $M_{n,n-d}(K)$ with respect to the 
Zariski topology and hence defines a $K$-manifold. Furthermore, we have a 
mapping $\text{LK}_r : \mathcal M_r \to \Grass(n,n-r)$ which is 
differentiable and whose differential can be computed as before.
